Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2022-02-21 Thesis/_book/thesis.tex
update exp design

@@

Prolific provides an international participant pool with the option to apply flexible pre-screening filters. In this study, we recruited participants who is fluent in English with at least \(10\) previous submissions and \(98\)\% approval rate in other Prolific studies for quality control. Further, balance sample across gender was imposed to prevent gender bias.

@@

Throughout the experiments, participants were requested to complete a short survey and evaluate 20 lineups on a website in an hour. Each lineup each consists of 19 null residual plots and 1 actual residual plot. Among the 20 lineups, there are 2 extremely east lineups used as attention checks which everyone should get correct. The short survey was intended to collect information about participant that might affect their ability in reading data plot including age, highest education, preferred pronoun and previous experience in similar study. For the evaluation of the lineup, participant first needed to select one or multiple most different data plot(s) from the lineup by clicking the corresponding image. Then, the reason for choosing the plots needed to be provided by selecting one of the given options - ``outlier(s)'', ``cluster(s)'', ``shape'' and ``other''. Explanations about these options were provided in the training page and the mouseover text. Table \ref{tab:lineupreason} gives the detailed explanations about these reasons. If the option ``other'' was selected, text input for the specified reason would be collected. Lastly, the degree of difference between their chosen data plot(s) and other data plots needed to be selected among \(5\) levels - ``not at all'', ``slightly'', ``moderately'', ``very'' and ``extremely''.

@@

The study website was powered by Flask \autocite{grinberg_flask_2018}, a web framework written in Python 3 \autocite{van_rossum_python_2009}, hosted on PythonAnywhere \autocite{pythonanywhere_pythonanywhere_2022}, a web hosting service provider. The website is built using a JavaScript \autocite{flanagan_javascript_2006} library jsPsych \autocite{de_leeuw_jspsych_2015} which is specialized in creating online behavioral experiments. One of the reasons we chose to use this library was it had a modularized but highly customizable template which could record participant's response time automatically. This is essential for us to confirm the quality of the data by checking exceptionally fast or slow responses. The first two pages of the website are the explanatory statement and the consent form. Participants needed to read the documentation and agree with the terms to advance to the short survey. With the completion of the survey, the next page is the training page, which contained instructions on how to evaluate the lineup, give the reason and choose the confidence level. Followed by the training page, there are 20 lineups each on a single page. At the end of the experiment, the participants would be redirected back to Prolific and waited for researchers' responses.

@@

Two models were used in both experiments, which were linear models with some degree of violations of classical assumptions. Residuals vs.~fitted values plots were used to diagnose the issues.

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(visage)}
\NormalTok{mod }\OtherTok{\textless{}{-}} \FunctionTok{cubic\_model}\NormalTok{(}\AttributeTok{a =} \SpecialCharTok{{-}}\DecValTok{1}\NormalTok{, }\AttributeTok{b =} \SpecialCharTok{{-}}\DecValTok{1}\NormalTok{, }\AttributeTok{c =} \DecValTok{1}\NormalTok{, }\AttributeTok{sigma =} \FloatTok{0.5}\NormalTok{, }
                   \AttributeTok{x =} \FunctionTok{rand\_uniform}\NormalTok{(}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{), }\AttributeTok{z =} \FunctionTok{rand\_uniform}\NormalTok{(}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\begin{verbatim}
##             y          x          z          e      .resid     .fitted
## 1  0.72328137  0.8783282 0.17687898  0.1539500  0.37649863  0.34678274
## 2 -0.01317346  0.9593166 0.30200008 -0.3526064 -0.22899266  0.21581920
## 3  0.10405716  0.7677855 0.72090641 -0.4481692  0.36279717 -0.25874001
## 4 -0.53007353  0.9238035 0.55802451 -0.8849474 -0.46221585 -0.06785768
## 5  0.33947802 -0.7784446 0.02191688  0.2307539 -0.04808729  0.38756531
\end{verbatim}

@@

@@

\begin{verbatim}
## [1]  0.05836465 -0.03566983 -0.09457480 -0.01126447  0.08314445
\end{verbatim}

@@

To obtain a measure of the impact of \(a\) and \(b\) on the residuals, we need to address several properties of the residuals. First, because \(\boldsymbol{R}_a\) is not an identity matrix, the residuals will have different variance and will be correlated with each other. This can be fixed by standardizing the residuals by their variance-covariance matrix. Second, the difference between \(\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b\) and \(\boldsymbol{0}\) could be negative, which is not ideal for comparison. Thus, the magnitude needs to be squared. Third, the measure needs to be a scalar. We could apply a weighted average operator \(\boldsymbol{W}\) on the transformed expected residuals to obtain a single numeric value. For simplicity, we set \(\boldsymbol{W}=n^{-1}\boldsymbol{1}\). This gives the effect size: \[ES=n^{-1}||\sigma^{-1}\boldsymbol{R}_a^{-\frac{1}{2}}\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b||^2=n^{-1}\sigma^{-2}||\boldsymbol{R}_a^{\frac{1}{2}}\boldsymbol{X}_b\boldsymbol{\beta}_b||^2\approx n^{-1}\sigma^{-2}||diag(\boldsymbol{R}_a)^{\frac{1}{2}}\boldsymbol{X}_b\boldsymbol{\beta}_b||^2.\]

The interpretation of this effect size is the impact of parameter \(a\) and \(b\) on the squared deviation of the standardized expected residual per observation. It is not directly related to the shape, or the pattern observed from the residual plot, but it is a reasonable approximation of the degree of the deviation from the null residuals under our cubic model setting.

\includegraphics{thesis_files/figure-latex/unnamed-chunk-5-1.pdf}

>>>>

\textbf{null data generate mechanism}

@@

Prolific provides an international participant pool with the option to apply flexible pre-screening filters. In this study, we recruited participants who was fluent in English with at least \(10\) previous submissions and \(98\)\% approval rate in other Prolific studies for quality control. Further, balance sample across gender was imposed to prevent gender bias.

@@

Throughout the experiments, participants were requested to complete a short survey and evaluate 20 lineups on a website in an hour. Each lineup consists of one actual residual plot and 19 null residual plots produced by plotting null residuals simulated from the residual rotation distributions. Among the 20 lineups, there are two extremely easy lineups used as attention checks which everyone should get correct. The short survey was intended to collect information about participant that might affect their ability in reading data plot including age, highest level of education, preferred pronoun and previous experience in similar study. For the evaluation of the lineup, participant first needed to select one or multiple most different data plot(s) from the lineup by clicking the corresponding image. Then, the reason for choosing the plots needed to be provided by selecting one of the given options - ``outlier(s)'', ``cluster(s)'', ``shape'' and ``other''. Explanations about these options were provided in the training page and the mouseover text. Table \ref{tab:lineupreason} gives the detailed explanations about these reasons. If the option ``other'' was selected, text input for the specified reason would be collected. Lastly, the degree of difference between their chosen data plot(s) and other data plots needed to be selected among \(5\) levels - ``not at all'', ``slightly'', ``moderately'', ``very'' and ``extremely''. Notably, if participants could not tell the difference between the data plots, there was an option to skip the evaluation of the lineup. However, to prevent participants abusing this option, warnings were given at the beginning of the study that skipping too many lineups might lead to rejection of submission since it demonstrated clear low-effort throughout the experiment.

@@

The study website was powered by Flask \autocite{grinberg_flask_2018}, a web framework written in Python 3 \autocite{van_rossum_python_2009}, hosted on PythonAnywhere \autocite{pythonanywhere_pythonanywhere_2022}, a web hosting service provider. Due to the limit of the storage on PythonAnywhere, lineup images needed for the website were hosted separately on Github Pages \autocite{gtihub_github_2022-1}, a static site hosting service provided by Github \autocite{gtihub_github_2022}, stored in private Github repositories. The uniform resource locator (URL) to lineup images were set to be unique random strings such that images can not be accessed by general approaches without knowing the correct URL. This avoided participants seeing the lineup beforehand. The front-end of the website was built using a JavaScript \autocite{flanagan_javascript_2006} library jsPsych \autocite{de_leeuw_jspsych_2015} which is specialized in creating online behavioral experiments. One of the reasons we chose to use this library was it had a modularized but highly customizable template which could record participant's response time automatically. This is essential for us to confirm the quality of the data by checking exceptionally fast or slow responses. Figure \ref{fig:DiagramExperiment} summarizes the experiment setup.

\begin{figure}
\centering
\includegraphics[width=4.6875in,height=4.6875in]{figures/experiment_tech.pdf}
\caption{Diagram of online experiment setup. The server-side of the study webiste used Flask as backend hosted on PythonAnywhere. And the client-side used jsPsych to run experiment. \label{fig:DiagramExperiment}}
\end{figure}

The first two pages of the website are the explanatory statement and the consent form. Participants needed to read the documents and agree with the terms to advance to the short survey. With the completion of the survey, the next page is the training page, which contained instructions on how to evaluate the lineup, give the reason and choose the confidence level. Followed by the training page, there are 20 lineups each on a single page. Figure \ref{fig:WebsiteLayout} illustrates the layout of the website. At the end of the experiment, the participants would be redirected back to Prolific and waited for researchers' responses.

\begin{figure}
\centering
\includegraphics[width=5.72917in,height=3.11458in]{figures/website.png}
\caption{Layout of the study website. Participants needed to choose the most different plots on the right and select their reasons and confidence levels on the left. \label{fig:WebsiteLayout}}
\end{figure}

@@

Two models were used in both experiments, which were linear models with some degree of violations of classical assumptions. Residual vs.~fitted values plot were used to present the model defects.

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(visage)}
\NormalTok{mod }\OtherTok{\textless{}{-}} \FunctionTok{cubic\_model}\NormalTok{(}\AttributeTok{a =} \SpecialCharTok{{-}}\DecValTok{3}\NormalTok{, }\AttributeTok{b =} \SpecialCharTok{{-}}\DecValTok{3}\NormalTok{, }\AttributeTok{c =} \DecValTok{1}\NormalTok{, }\AttributeTok{sigma =} \FloatTok{0.5}\NormalTok{, }
                   \AttributeTok{x =} \FunctionTok{rand\_uniform}\NormalTok{(}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{), }\AttributeTok{z =} \FunctionTok{rand\_uniform}\NormalTok{(}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\begin{verbatim}
##             y          x          z          e      .resid   .fitted
## 1 -2.24847020  0.8783282 0.17687898  0.1539500  0.65761568 -2.906086
## 2 -3.85694079  0.9593166 0.30200008 -0.3526064 -0.45596841 -3.400972
## 3 -3.76887401  0.7677855 0.72090641 -0.4481692  0.56311185 -4.331986
## 4 -4.78398171  0.9238035 0.55802451 -0.8849474 -0.68606690 -4.097915
## 5  0.06998178 -0.7784446 0.02191688  0.2307539 -0.07869221  0.148674
\end{verbatim}

@@

Lineup can be produced by using the methods \texttt{mod\$gen\_lineup()} and \texttt{mod\$plot\_lineup()}. Method \texttt{mod\$gen\_lineup} takes the number of realizations \texttt{n} and the number of plots in a lineup \texttt{k} as inputs. And the method \texttt{mod\$plot\_lineup()} has the same user interface as \texttt{mod\$plot()}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{mod}\SpecialCharTok{$}\FunctionTok{plot\_lineup}\NormalTok{(mod}\SpecialCharTok{$}\FunctionTok{gen\_lineup}\NormalTok{(}\AttributeTok{n =} \DecValTok{50}\NormalTok{, }\AttributeTok{k =} \DecValTok{20}\NormalTok{), }\AttributeTok{type =} \StringTok{"resid"}\NormalTok{, }\AttributeTok{size =} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{thesis_files/figure-latex/unnamed-chunk-4-1.pdf}

@@

\begin{verbatim}
## [1]  1.6431906  1.2501862 -1.9594998 -1.4124888  0.4786119
\end{verbatim}

@@

To obtain a measure of the impact of \(a\) and \(b\) on the residuals adjusted for variance and covariance, we need to address several properties of the residuals. First, the variance of the residuals \(\sigma^2\boldsymbol{R}_a\) is not an identity matrix. This can be fixed by standardizing the residuals by their variance-covariance matrix. Second, the difference between \(\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b\) and \(\boldsymbol{0}\) could be negative, which is not ideal for comparison. Thus, the magnitude needs to be squared. Third, the measure needs to be a scalar. We could apply a weighted average operator \(\boldsymbol{W}\) on the transformed expected residuals to obtain a single numeric value. For simplicity, we set \(\boldsymbol{W}=n^{-1}\boldsymbol{1}\). Considering the high time complexity of computing the square root of \(\boldsymbol{R}_a\), off-diagonal elements of \(\boldsymbol{R}_a\) are set to be zeros. This gives the effect size: \[ES=n^{-1}||\sigma^{-1}\boldsymbol{R}_a^{-\frac{1}{2}}\boldsymbol{R}_a\boldsymbol{X}_b\boldsymbol{\beta}_b||^2=n^{-1}\sigma^{-2}||\boldsymbol{R}_a^{\frac{1}{2}}\boldsymbol{X}_b\boldsymbol{\beta}_b||^2\approx n^{-1}\sigma^{-2}||diag(\boldsymbol{R}_a)^{\frac{1}{2}}\boldsymbol{X}_b\boldsymbol{\beta}_b||^2,\] where \(diag(\boldsymbol{R}_a)\) is the diagonal matrix constructed from the diagnoal element of \(\boldsymbol{R}_a\).

The interpretation of this effect size is the impact of parameter \(a\) and \(b\) on the squared deviation of the standardized expected residual per observation. It is not directly related to the shape, or the pattern human observed from the residual plot, but it is a reasonable approximation of the degree of the visual deviation from the null residuals under our cubic model setting. Figure \ref{fig:effectsize} shows four residual plots with different effect sizes. As effect sizes increases, the strength of the signal become stronger.

\begin{figure}
\centering
\includegraphics{thesis_files/figure-latex/effectsize-1.pdf}
\caption{\label{fig:effectsize}Cubic model residual plots under different effect sizes.}
\end{figure}
